Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2018-11-01 manuscript/Manuscript_Claire.Rmd
Ready for resubmission - typos fixed

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplmentary Material). Papers with 3, 4 or $\ge5$ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3 author) and larger (5+ authors) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconcious selection of same-gender collaborators by men, by women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these uncontrolled artefacts to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals relative to specialised journals; restricting the data by country yielded similar estimates of $\alpha'$; and we used a simple model to show that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers preferentially select same-gendered collaborators, although it is difficult to ascertain what proportion of people show such a preference, or how much the strength of the preference varies between individuals. We also note that even in a world in which collaboration was completely random with respect to gender, a high proportion of individual researchers would have entirely same-gendered collaborators by chance alone (especially in gender-biased disciplines); thus, individuals who only have same-gendered co-authors are not necessarily doing anything differently from people with gender-balanced co-authors.

@@

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? We beleive that the answers to these questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since the results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to gender differences in confidence, or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality (which is contentious; [@Garfield_2006]), our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

@@

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to infer each author's gender from their given name using computational methods. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations in the interests of objectivity. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we only kept papers with publicaiton dates falling in two one-year time periods, namely 0-1 or 10-11 years prior to the collection date of the PubMed data (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. Detailed sample size information is given in S1 Table.

@@

To model the relationship between $\alpha'$ and the number of authors on the paper, we used a meta-regression model implemented in the R package `brms` [@burkner_2016]. The model incorporated the standard error associated with easch estimate of $\alpha'$, had author number as a fixed effect, and journal as a random intercept (to control for repeated measures of each journal). We also fit a random slope of author number within journal, thereby allowing the response to author number to vary between journals. We used the default (weak) priors. The full output of this model can be viewed in the Online Supplementary Material.

@@

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine if and how this effect should alter our inferences. For simplicity, our calculations assume there are only two career stages ('early-career' and 'established'), though we expect that the general conclusions would also apply to a multi-tier career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters in our simple model. These parameters are: the gender ratio among early-career researchers (x-axis of Figure 6), the gender ratio among established researchers (y-axis of Figure 6), the frequency of within- versus between career stage collaborator pairs (rows in Figure 6), and lastly the frequency of within-stage collaborations that are between two early-career researchers as opposed to two late-career reserachers (columns in Figure 6). When these four parameters are specified, one can easily calculate the relative frequencies of collaborator pairs that involve two men, two women, or a man and a woman. In short, if we have specified the frequency of women at both career stages, as well as the frequency of the three possible types of collaboration with respect to career stage (early-early, early-established, and established-established), then we can calculate the frequency of collaborators pairs comprising two women, or a woman and a man, and thus find $\alpha$ (see the Online Supplementary Material for the annotated R code).

@@

S6 Fig. Histogram showing the estimated degree to which $\alpha'$ is inflated by inter-country differences in author gender ratio, across the 285 journals for which we had adequate data after restricting the analysis by country. The average inflation in $\alpha'$ is negligible, suggesting that Wahlund effects resulting from inter-country differences have a neglible effect on our estimates of gender homophily.

@@

S2 Table. Number of journals showing statistically significant homophily or heterophily, in two one-year periods. The significance threshold was p < 0.05, and p-values were adjused using Benjamini-Hochberg false discovery rate correction. Note that the power of our test is lower for the 2005-2006 data because fewer papers were recovered per journal: thus, it is not meaningful to compare the % significant journals (i.e. 11% vs 24%) between the two time periods.

>>>>

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplementary Material). Papers with 3, 4 or $\ge5$ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers. One possible explanation for this finding is that 2-authors papers might be more likely to have an author list that is evenly split between career stages (e.g. a postgraduate student and their supervisor), increasing the chance that the authors are mixed gender (see Figure 6). The result also suggests that the processes responsible for gender homophily are similar in small (e.g. 3 author) and larger (5+ authors) collaborations (and across disciplines where small versus large collaborations are the norm).

@@

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconscious selection of same-gender collaborators by men, by women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these uncontrolled artefacts to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals relative to specialised journals; restricting the data by country yielded similar estimates of $\alpha'$; and we used a simple model to show that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers preferentially select same-gendered collaborators, although it is difficult to ascertain what proportion of people show such a preference, or how much the strength of the preference varies between individuals. We also note that even in a world in which collaboration was completely random with respect to gender, a high proportion of individual researchers would have entirely same-gendered collaborators by chance alone (especially in gender-biased disciplines); thus, individuals who only have same-gendered co-authors are not necessarily doing anything differently from people with gender-balanced co-authors.

@@

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? We believe that the answers to these questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since the results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to gender differences in confidence, or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality (which is contentious; [@Garfield_2006]), our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

@@

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to infer each author's gender from their given name using computational methods. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations in the interests of objectivity. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we only kept papers with publication dates falling in two one-year time periods, namely 0-1 or 10-11 years prior to the collection date of the PubMed data (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. Detailed sample size information is given in S1 Table.

@@

To model the relationship between $\alpha'$ and the number of authors on the paper, we used a meta-regression model implemented in the R package `brms` [@burkner_2016]. The model incorporated the standard error associated with each estimate of $\alpha'$, had author number as a fixed effect, and journal as a random intercept (to control for repeated measures of each journal). We also fit a random slope of author number within journal, thereby allowing the response to author number to vary between journals. We used the default (weak) priors. The full output of this model can be viewed in the Online Supplementary Material.

@@

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine if and how this effect should alter our inferences. For simplicity, our calculations assume there are only two career stages ('early-career' and 'established'), though we expect that the general conclusions would also apply to a multi-tier career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters in our simple model. These parameters are: the gender ratio among early-career researchers (x-axis of Figure 6), the gender ratio among established researchers (y-axis of Figure 6), the frequency of within- versus between career stage collaborator pairs (rows in Figure 6), and lastly the frequency of within-stage collaborations that are between two early-career researchers as opposed to two late-career researchers (columns in Figure 6). When these four parameters are specified, one can easily calculate the relative frequencies of collaborator pairs that involve two men, two women, or a man and a woman. In short, if we have specified the frequency of women at both career stages, as well as the frequency of the three possible types of collaboration with respect to career stage (early-early, early-established, and established-established), then we can calculate the frequency of collaborators pairs comprising two women, or a woman and a man, and thus find $\alpha$ (see the Online Supplementary Material for the annotated R code).

@@

S6 Fig. Histogram showing the estimated degree to which $\alpha'$ is inflated by inter-country differences in author gender ratio, across the 285 journals for which we had adequate data after restricting the analysis by country. The average inflation in $\alpha'$ is negligible, suggesting that Wahlund effects resulting from inter-country differences have a negligible effect on our estimates of gender homophily.

@@

S2 Table. Number of journals showing statistically significant homophily or heterophily, in two one-year periods. The significance threshold was p < 0.05, and p-values were adjusted using Benjamini-Hochberg false discovery rate correction. Note that the power of our test is lower for the 2005-2006 data because fewer papers were recovered per journal: thus, it is not meaningful to compare the % significant journals (i.e. 11% vs 24%) between the two time periods.

### 2018-11-01 manuscript/Manuscript_Claire.tex
Ready for resubmission - typos fixed

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplmentary Material). Papers with 3, 4 or \(\ge5\) authors had
essentially identical average \(\alpha'\) values. The variance in
\(\alpha'\) across journals was also a little higher for 2-authors
papers compared to the remainder (\autoref{author_number}), though part
of this variance is due to the reduced sample size (in terms of number
of authors) for the 2-author papers. One possible explanation for this
finding is that 2-authors papers might be more likely to have an author
list that is evenly split between career stages (e.g.~a postgraduate
student and their supervisor), increasing the chance that the authors
are mixed gender (see Figure 6). The result also suggests that the
processes responsible for gender homophily are similar in small (e.g.~3
author) and larger (5+ authors) collaborations (and across disciplines
where small versus large collaborations are the norm).

@@

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconcious selection of same-gender collaborators by men, by women, or
both -- our results could be partly explained by uncontrolled Wahlund
effects. However, we suspect the contribution of these uncontrolled
artefacts to be minor, for four reasons: we found positive \(\alpha'\)
after controlling for three obvious sources of Wahlund effect; there was
no inflation of \(\alpha'\) in highly multidisciplinary journals
relative to specialised journals; restricting the data by country
yielded similar estimates of \(\alpha'\); and we used a simple model to
show that differences in gender ratio between career stages are unlikely
to fully explain our results. On balance, we believe the data suggest
that it is likely that some researchers preferentially select
same-gendered collaborators, although it is difficult to ascertain what
proportion of people show such a preference, or how much the strength of
the preference varies between individuals. We also note that even in a
world in which collaboration was completely random with respect to
gender, a high proportion of individual researchers would have entirely
same-gendered collaborators by chance alone (especially in gender-biased
disciplines); thus, individuals who only have same-gendered co-authors
are not necessarily doing anything differently from people with
gender-balanced co-authors.

@@

Our study begs two questions: what causes gender homophily in science,
and are our results cause for concern? We beleive that the answers to
these questions are closely related. For example, some of the homophily
we observed might be caused by women seeking to avoid harassment or
sexism from men {[}38{]}, which would clearly be concerning.
Additionally, Sheltzer and Smith {[}66{]} concluded that `elite' male
academics (defined as recipients of major honours) have a higher
proportion of male students and postdocs than non-elite male academics.
This finding could contribute to the homophily we observed, and is cause
for concern since the results might reflect discrimination against women
during hiring {[}20{]}, or avoidance by women of elite research groups
(e.g.~due to gender differences in confidence, or a perception that some
groups are sexist). We also found a little evidence that gender
homophily is detrimental to research quality, in that high-impact
journals tended to have weaker homophily. Assuming that papers published
in high-impact journals are of higher average quality (which is
contentious; {[}67{]}), our results provide non-experimental support for
the hypothesis that mixed-gender teams produce better research than
single-gender teams {[}42--48{]}. Another issue is that if many
collaborations are between established researchers, there will be an
excess of male-male collaborations in fields where women in senior
positions are rare; some of the observed homophily might therefore
reflect the elevated gender gap among senior researchers.

@@

We used the dataset of PubMed author lists from Holman et al. {[}5{]}.
Briefly, that dataset was created by downloading every article indexed
on PubMed and attempting to infer each author's gender from their given
name using computational methods. Each journal was assigned to one of
107 scientific disciplines, using PubMed's journal categorisations in
the interests of objectivity. Because the present study focuses on
co-authorship, all single-author papers were discarded. We also
discarded all papers for which we could not determine the gender of
every author with \({\ge}95\%\) certainty, in order to simplify the
statistical analysis. To mitigate Wahlund effects caused by variation in
the gender ratio of researchers over time (see below), we only kept
papers with publicaiton dates falling in two one-year time periods,
namely 0-1 or 10-11 years prior to the collection date of the PubMed
data (i.e.~20\textsuperscript{th} August 2016). Lastly, we excluded
journals with fewer than 50 suitable papers. Detailed sample size
information is given in S1 Table.

@@

To model the relationship between \(\alpha'\) and the number of authors
on the paper, we used a meta-regression model implemented in the R
package \texttt{brms} {[}75{]}. The model incorporated the standard
error associated with easch estimate of \(\alpha'\), had author number
as a fixed effect, and journal as a random intercept (to control for
repeated measures of each journal). We also fit a random slope of author
number within journal, thereby allowing the response to author number to
vary between journals. We used the default (weak) priors. The full
output of this model can be viewed in the Online Supplementary Material.

@@

We can think of no tractable method of controlling for this issue using
our dataset, which contains no information on career stage. Therefore,
we instead decided to derive theoretical expectations for \(\alpha\)
when there is a difference in gender ratio across career stages, in
order to determine if and how this effect should alter our inferences.
For simplicity, our calculations assume there are only two career stages
(`early-career' and `established'), though we expect that the general
conclusions would also apply to a multi-tier career ladder. Under the
null model that gender has no causal effect on collaboration, we
calculated \(\alpha\) for various combinations of the four free
parameters in our simple model. These parameters are: the gender ratio
among early-career researchers (x-axis of Figure 6), the gender ratio
among established researchers (y-axis of Figure 6), the frequency of
within- versus between career stage collaborator pairs (rows in Figure
6), and lastly the frequency of within-stage collaborations that are
between two early-career researchers as opposed to two late-career
reserachers (columns in Figure 6). When these four parameters are
specified, one can easily calculate the relative frequencies of
collaborator pairs that involve two men, two women, or a man and a
woman. In short, if we have specified the frequency of women at both
career stages, as well as the frequency of the three possible types of
collaboration with respect to career stage (early-early,
early-established, and established-established), then we can calculate
the frequency of collaborators pairs comprising two women, or a woman
and a man, and thus find \(\alpha\) (see the Online Supplementary
Material for the annotated R code).

@@

S6 Fig. Histogram showing the estimated degree to which \(\alpha'\) is
inflated by inter-country differences in author gender ratio, across the
285 journals for which we had adequate data after restricting the
analysis by country. The average inflation in \(\alpha'\) is negligible,
suggesting that Wahlund effects resulting from inter-country differences
have a neglible effect on our estimates of gender homophily.

@@

S2 Table. Number of journals showing statistically significant homophily
or heterophily, in two one-year periods. The significance threshold was
p \textless{} 0.05, and p-values were adjused using Benjamini-Hochberg
false discovery rate correction. Note that the power of our test is
lower for the 2005-2006 data because fewer papers were recovered per
journal: thus, it is not meaningful to compare the \% significant
journals (i.e.~11\% vs 24\%) between the two time periods.

>>>>

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplementary Material). Papers with 3, 4 or \(\ge5\) authors had
essentially identical average \(\alpha'\) values. The variance in
\(\alpha'\) across journals was also a little higher for 2-authors
papers compared to the remainder (\autoref{author_number}), though part
of this variance is due to the reduced sample size (in terms of number
of authors) for the 2-author papers. One possible explanation for this
finding is that 2-authors papers might be more likely to have an author
list that is evenly split between career stages (e.g.~a postgraduate
student and their supervisor), increasing the chance that the authors
are mixed gender (see Figure 6). The result also suggests that the
processes responsible for gender homophily are similar in small (e.g.~3
author) and larger (5+ authors) collaborations (and across disciplines
where small versus large collaborations are the norm).

@@

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconscious selection of same-gender collaborators by men, by women, or
both -- our results could be partly explained by uncontrolled Wahlund
effects. However, we suspect the contribution of these uncontrolled
artefacts to be minor, for four reasons: we found positive \(\alpha'\)
after controlling for three obvious sources of Wahlund effect; there was
no inflation of \(\alpha'\) in highly multidisciplinary journals
relative to specialised journals; restricting the data by country
yielded similar estimates of \(\alpha'\); and we used a simple model to
show that differences in gender ratio between career stages are unlikely
to fully explain our results. On balance, we believe the data suggest
that it is likely that some researchers preferentially select
same-gendered collaborators, although it is difficult to ascertain what
proportion of people show such a preference, or how much the strength of
the preference varies between individuals. We also note that even in a
world in which collaboration was completely random with respect to
gender, a high proportion of individual researchers would have entirely
same-gendered collaborators by chance alone (especially in gender-biased
disciplines); thus, individuals who only have same-gendered co-authors
are not necessarily doing anything differently from people with
gender-balanced co-authors.

@@

Our study begs two questions: what causes gender homophily in science,
and are our results cause for concern? We believe that the answers to
these questions are closely related. For example, some of the homophily
we observed might be caused by women seeking to avoid harassment or
sexism from men {[}38{]}, which would clearly be concerning.
Additionally, Sheltzer and Smith {[}66{]} concluded that `elite' male
academics (defined as recipients of major honours) have a higher
proportion of male students and postdocs than non-elite male academics.
This finding could contribute to the homophily we observed, and is cause
for concern since the results might reflect discrimination against women
during hiring {[}20{]}, or avoidance by women of elite research groups
(e.g.~due to gender differences in confidence, or a perception that some
groups are sexist). We also found a little evidence that gender
homophily is detrimental to research quality, in that high-impact
journals tended to have weaker homophily. Assuming that papers published
in high-impact journals are of higher average quality (which is
contentious; {[}67{]}), our results provide non-experimental support for
the hypothesis that mixed-gender teams produce better research than
single-gender teams {[}42--48{]}. Another issue is that if many
collaborations are between established researchers, there will be an
excess of male-male collaborations in fields where women in senior
positions are rare; some of the observed homophily might therefore
reflect the elevated gender gap among senior researchers.

@@

We used the dataset of PubMed author lists from Holman et al. {[}5{]}.
Briefly, that dataset was created by downloading every article indexed
on PubMed and attempting to infer each author's gender from their given
name using computational methods. Each journal was assigned to one of
107 scientific disciplines, using PubMed's journal categorisations in
the interests of objectivity. Because the present study focuses on
co-authorship, all single-author papers were discarded. We also
discarded all papers for which we could not determine the gender of
every author with \({\ge}95\%\) certainty, in order to simplify the
statistical analysis. To mitigate Wahlund effects caused by variation in
the gender ratio of researchers over time (see below), we only kept
papers with publication dates falling in two one-year time periods,
namely 0-1 or 10-11 years prior to the collection date of the PubMed
data (i.e.~20\textsuperscript{th} August 2016). Lastly, we excluded
journals with fewer than 50 suitable papers. Detailed sample size
information is given in S1 Table.

@@

To model the relationship between \(\alpha'\) and the number of authors
on the paper, we used a meta-regression model implemented in the R
package \texttt{brms} {[}75{]}. The model incorporated the standard
error associated with each estimate of \(\alpha'\), had author number as
a fixed effect, and journal as a random intercept (to control for
repeated measures of each journal). We also fit a random slope of author
number within journal, thereby allowing the response to author number to
vary between journals. We used the default (weak) priors. The full
output of this model can be viewed in the Online Supplementary Material.

@@

We can think of no tractable method of controlling for this issue using
our dataset, which contains no information on career stage. Therefore,
we instead decided to derive theoretical expectations for \(\alpha\)
when there is a difference in gender ratio across career stages, in
order to determine if and how this effect should alter our inferences.
For simplicity, our calculations assume there are only two career stages
(`early-career' and `established'), though we expect that the general
conclusions would also apply to a multi-tier career ladder. Under the
null model that gender has no causal effect on collaboration, we
calculated \(\alpha\) for various combinations of the four free
parameters in our simple model. These parameters are: the gender ratio
among early-career researchers (x-axis of Figure 6), the gender ratio
among established researchers (y-axis of Figure 6), the frequency of
within- versus between career stage collaborator pairs (rows in Figure
6), and lastly the frequency of within-stage collaborations that are
between two early-career researchers as opposed to two late-career
researchers (columns in Figure 6). When these four parameters are
specified, one can easily calculate the relative frequencies of
collaborator pairs that involve two men, two women, or a man and a
woman. In short, if we have specified the frequency of women at both
career stages, as well as the frequency of the three possible types of
collaboration with respect to career stage (early-early,
early-established, and established-established), then we can calculate
the frequency of collaborators pairs comprising two women, or a woman
and a man, and thus find \(\alpha\) (see the Online Supplementary
Material for the annotated R code).

@@

S6 Fig. Histogram showing the estimated degree to which \(\alpha'\) is
inflated by inter-country differences in author gender ratio, across the
285 journals for which we had adequate data after restricting the
analysis by country. The average inflation in \(\alpha'\) is negligible,
suggesting that Wahlund effects resulting from inter-country differences
have a negligible effect on our estimates of gender homophily.

@@

S2 Table. Number of journals showing statistically significant homophily
or heterophily, in two one-year periods. The significance threshold was
p \textless{} 0.05, and p-values were adjusted using Benjamini-Hochberg
false discovery rate correction. Note that the power of our test is
lower for the 2005-2006 data because fewer papers were recovered per
journal: thus, it is not meaningful to compare the \% significant
journals (i.e.~11\% vs 24\%) between the two time periods.
